The role of cross-linguistic lexical similarity on bilingual word acquisition

### 2022-11-17 manuscript.tex
Update lab notes

\begin{document}
\maketitle
\begin{abstract}
Bilinguals face the challenging task of learning words from languages
with overlapping phonologies. Floccia et al.~(2018) reported larger
vocabulary sizes for 24-month-old bilinguals that were learning
languages that shared a greater amount of cognates (e.g.,
English-Dutch). The mechanisms underlying this effect remain unknown. We
explore two compatible scenarios. First, we test whether cognates are
learnt earlier than non-cognates. This would account for the difference
in vocabulary size associated to the amount of shared cognates across
languages. Second, we explore the possibility that the word-forms of one
language interact with those form the other language, scaffolding the
acquisition of their translation equivalents when their phonologies
overlap. This mechanism, in line with the parallel activation account of
bilingual speech perception, would provide a plausible explanation to
why cognates are acquired ealier by bilinguals. We developed an online
tool to collect parental reports of receptive and productive
vocabularies from children learning Catalan and/or Spanish, and present
data on receptive and productive vocabulary of bilingual toddlers aged
10 to 36 months.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable]}{\end{tcolorbox}}\fi

\setstretch{1.25}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The foundations of word learning are in place early in age: before the
end of their first year of life infants start directing their gaze to
some objects when hearing their labels, according to both experimental
data (Bergelson \& Swingley, 2012, 2015; Jusczyk \& Aslin, 1995; Tincoff
\& Jusczyk, 1999) and parental reports (e.g., Fenson et al., 2007).
During the last half of their second year, they acquire new words at an
increasingly fast rate (Bergelson, 2020; Bloom, 2002; Fenson et al.,
1994; Goldfield \& Reznick, 1990; Mayor \& Plunkett, 2011). These early
stages of lexical acquisition are characterised by substantial variation
across children, reflected, for instance, on the variation on the number
of words they know (i.e.~vocabulary size, Fenson et al., 1994; Frank et
al., 2021) or on the proportion of those words that fall into the
category of nouns, as opposed to verbs, adjectives, or function words
(e.g., Nelson, 1973, p. bates1994developmental).

@@

There is evidence that bilinguals know, on average, less words in each
of their languages than monolinguals, but also that both groups know a
similar amount of words when the two languages are aggregated. For
example, Hoff et al. (2012) found that English-Spanish bilingual
toddlers in South Florida knew less words in English than monolinguals,
who only learnt English. Both groups knew a similar total amount of
words when both English and Spanish vocabularies were counted together,
pointing to the importance of collecting data on both languages when
assessing bilinguals' communicative development. Other studies have
provided converging evidence that bilinguals know a similar--or even
larger--number of words than monolinguals, only when the languages are
aggregated (Gonzalez-Barrero et al., 2020; Oller \& Eilers, 2002; J.
Patterson et al., 2004; J. L. Patterson, 2004; Pearson et al., 1993;
Pearson \& Fernández, 1994; Petitto et al., 2001; Smithson et al.,
2014). While these studies have mostly relied on samples of bilingual
children learning two relatively distant languages, as it is the case of
English and Spanish, it is unclear whether children learning
typologically more similar languages also know less words in each of
their languages than monolinguals. What role could linguistic distance
play during early vocabulary growth?

@@

Floccia et al.~pointed to \emph{parallel activation} as the main
mechanism underpinning their results. The parallel activation hypothesis
suggests that bilinguals activate both languages simultaneously during
speech production or comprehension, and that this phenomenon is the
result of the activation of lexical representations in both languages,
even when only one is in use during production (Costa et al., 2000;
Hoshino \& Kroll, 2008) or comprehension (Spivey \& Marian, 1999;
Thierry \& Wu, 2007). One of the clearest examples of parallel
activation was provided by Costa et al. (2000). In this study,
Catalan-Spanish monolingual and bilingual adults were asked to name
pictures of common objects in Spanish. In half of the trials, the object
labels were cognates in Spanish and Catalan (\emph{árbol}-\emph{arbre},
translations of \emph{tree}), whereas in the other half of the trials
labels were non-cognates (\emph{mesa}-\emph{taula}, translations of
\emph{table}). Bilinguals named cognate pictures faster than non-cognate
pictures, even after adjusting for the lexical frequency of the items.
Importantly, Spanish monolinguals--unfamiliar with the Catalan
translations of the Spanish words they uttered--did not show this
effect. These results suggest that bilinguals' Catalan phonology was
activated during the production of Spanish words, facilitating the
naming of cognate pictures. Several subsequent studies have also
provided similar evidence in comprehension in children (Poulin-Dubois et
al., 2013; e.g., Von Holzen \& Mani, 2012). Parallel activation is
therefore a plausible mechanism to account for Floccia et al.'s results:
cognates increase the amount of activation in both languages,
facilitating word acquisition, and ultimately leading to children
learning language pairs with a larger proportion of cognates are
predicted to show larger vocabulary sizes. A missing piece in this
account is how increase parallel activation might facilitate the
acquisition new words.

In this study, we hypothesise that cognates are acquired earlier in age
than non-cognates, which might explain why children learning languages
that share many cognates show larger vocabulary sizes. Evidence
supporting this claim is scarce. Bosch \& Ramon-Casas (2014) used
vocabulary parental reports (152 lexical items) from 48 Catalan-Spanish
bilinguals aged 18 months, and found that cognates represented a larger
proportion of participant's vocabulary than non-cognates. Schelletter
(2002) reported a longitudinal single case of one English-German
bilingual who produced cognates earlier than non-cognates, on average.
The low sample size in these two studies makes it challenging to draw
strong conclusions about the effect of cognateness on vocabulary growth.
On the other hand, Floccia et al.'s estimates are statically more
reliable given their (much larger) sample size, but their study was not
aimed at testing the effect of cognateness on age of acquisition
directly. In their discussion the authors state the following (pp.~70):

@@

As Mitchell et al. (2022) note in their discussion, two competing
hypotheses remain untested. The facilitative effect of cognateness can
be explained by, at least, two alternative, arguably incompatible,
mechanisms. On the one hand, cognateness might operate as a facilitator
of the acquisition of both forms of a translation pair, with every
exposure to either form activating the other, and therefore
strengthening the association between both and their referent. If this
is the case, then cognateness should increase the probability of
acquisition of the forms in both languages equally.

Another possibility is that cognateness plays a role in the acquisition
of a translation pair only when at least one of the word-forms has been
acquired. This account departs from the premise that, if the child is
not previously familiar with one of the word forms, the acquisition of
its translation in the other language cannot be facilitated by
cognateness. This implies that, if cognateness facilitates the
acquisition of a word form, it can only do so after the child has
acquired one of the word forms of the translation pair. Given that
children are more likely to acquire words from languages to which they
are exposed more often (Cattani et al., 2014; David \& Wei, 2008;
Thordardottir et al., 2006), the acquisition of words in the language of
lower exposure should, on average, be more susceptible to the effect of
cognateness.

In line with this second hypothesis, the vocabulary size associated with
linguistic similarity that Floccia et al.~reported was larger in the
additional language vocabulary than in English vocabulary. Most
participants in their sample were English-dominant, meaning that their
relative amount of exposure to English was larger than in the additional
language. Therefore, participants may have, on average, learnt the
English word-form of translation equivalents earlier than the word-form
in the additional language. If this is the case, then the acquisition of
English words by English-dominant participants would rarely benefit from
their cognate status (the other word-form is not available yet), while
the acquisition of words in the additional language would benefit from
their phonological similarity with the (available) English form.

In this study, we investigated the role of cognateness on the
acquisition of translation equivalent, and whether such effect is
conditional to the amount of time the child is exposed to the
corresponding language of each member of the translation pair. Using an
online vocabulary checklist--designed specifically for this study--we
collected data from a sample of children aged 10 to 35 months learning
Catalan and/or Spanish, with varying degrees of exposure to each
language. We then adopted an Item Response Theory (\emph{IRT}) approach
to model the probability of participants being reported by their parents
to understand or understand \emph{and} say each word in the checklist,
conditional to its cognate status in Catalan and Spanish and
participants' degree of exposure to the corresponding language, while
adjusting for other indices of word difficulty--lexical frequency and
length (phonemes)-- and participant skill--age (see Kachergis et al.,
2022 for a similar approach).

\hypertarget{methods}{%
\section{Methods}\label{methods}}

Data and materials are available at OSF \url{https://osf.io/hy984/}, and
code is available at GitHub
\url{https://github.com/gongcastro/trajectories}.

\hypertarget{participants}{%
\subsection{Participants}\label{participants}}

@@

\hypertarget{questionnaire}{%
\subsection{Questionnaire}\label{questionnaire}}

The questionnaire was implemented on-line using the formR platform
(Arslan et al., 2020), and was structured in three blocks: a (1)
language questionnaire, a (2) demographic survey, and a (3) Catalan and
a Spanish vocabulary checklists. Vocabulary checklists followed a
similar structure as the Oxford Communicative Developmental Inventory
(Hamilton et al., 2000) and consisted in two lists of words: one in
Catalan and one in Spanish. The Catalan inventory contained 798 items
and the Spanish inventory contained 802. Items in one language were
translation equivalents of the items in the other language (e.g.,
whenever \emph{gos} {[}dog{]} was included in the Catalan inventory, the
word \emph{perro} was included in the Spanish inventory), roughly
following a one-to-one mapping. In case two translation equivalents were
possible for a given word, both were included as separate items (e.g.,
Catalan \emph{acabar} {[}\emph{to finish}{]} and Spanish \emph{acabar}
and \emph{terminar}), or merged them into a single item (e.g., Spanish
\emph{mono} {[}\emph{monkey}{]} and Catalan \emph{mono/mico}). We
included items from a diverse sample of 27 semantic/functional
categories.

For each word in the vocabulary checklists, we asked parents to report
whether their child was able to understand it, understand \emph{and} say
it, or did not understand or say it (checked out by default). Some
families filled a long version of the vocabulary checklists (803
translation equivalents; 798 items in Catalan, 802 items in Spanish),
while others filled a shorter version (\textasciitilde400 translation
equivalents, \textasciitilde400 items in Catalan, \textasciitilde400
items in Spanish). These last families were randomly allocated into one
of four different subsets of the complete list of items. These lists
were designed so that each contained a representative sub-sample of the
items from the complete list. Semantic/functional categories with less
than 16 items--thus resulting in less than four items after dividing it
in four lists--were not divided in the short version of the
questionnaire: all of their items were included in the four lists.
Another subset of items that were part of the trial lists of some
experiments in the lab were also included in all versions. For the
analyses in the present study, we considered responses to words
corresponding to nouns, verbs, and adjectives (Fourtassi et al., 2020).
We excluded multi-word items (e.g., \emph{barrita de cereales} {[}cereal
bar{]}) and items that included more than one word-form (e.g.,
\emph{mono / mico}).

We extracted four variables of interest for each word-form: (1) Word
lexical frequency (\(Frequency\)): lexical frequency of the word in its
corresponding language, expressed as Zipf scores (Van Heuven et al.,
2014; Zipf, 1949). This variable ranges from 0 to 7, and follows and
approximates a normal distribution, with most values in a corpus ranging
from 3 to 5 points. Lexical frequencies were extracted from an adult
subtitle-based database: SUBTLEX-CAT (Boada et al., 2020) for Catalan
words and SUBTLEX-ESP (Cuetos et al., 2012) for Spanish words. Using
child-based estimates of lexical frequency like CHILDES was not possible
due to the low number of Catalan participants and tokens in the
available corpora. Responses to words with missing lexical frequency
score were excluded from analyses (1.22\% of the total number of items).
(2) Word length (\(Phonemes\)): number of phonemes in the phonological
transcription of the word-form in International Phonological association
format. (3) Phonological similarity (\(Levenshtein\)): normalised
Levenshtein similarity between the word-form and its translation in the
other language (Levenshtein et al., 1966). This score is calculated by
first calculating the Levenshtein distance between the two
transcriptions (number of insertions, deletions or replacements needed
for the shortest transcription to become identical to the longer
transcription), then dividing the resulting value by the length of the
longest transcription, and finally subtracting this value from 1. This
results in a proportion that indicates how much the two phonological
transcriptions of the translation equivalent are similar to each other,
ranging from 0\% (no similarity at all) to 100\% (both transcriptions
are identical) (see Floccia et al., 2018; Fourtassi et al., 2020; Laing,
2022 for similar approaches). (4) Participant age: number of months
elapsed between participants' birth date and questionnaire completion.
Finally, (5) participant degree of exposure to the words' language
(\(DoE\)): percentage of exposure to the language the word belonged to.
For example, for a participant with 90\% exposure to Catalan, and 10\%
to Spanish, the DoE of the Catalan word \emph{taula} would be 90\%, and
the DoE for the Spanish word \emph{mesa} would be 10\%.
Table~\ref{tbl-item-summary} shows a summary of the distribution of the
word-level predictors across languages in and questionnaire versions.

\hypertarget{tbl-item-summary}{}
\begin{longtable}{lrrrrrrrrrr}
\caption{\label{tbl-item-summary}Lexical properties of the word-forms and translationpairs included in
the vocabulary checklist and our analyses. We report the number of
items, mean (M), standard deviation (SD), and range of the lexical
frequency (epxressed in Zipf scores), length (number of phonemes) of the
word forms, and the phonological similarity (normalised Levenshtein
similarity) with their translation equivalent. Statistics are reported
separately for Catalan and Spanish, and for the different vocabulary
checklists that participants were allocated to (A, B, C, D). }\tabularnewline

\toprule
 &  & \multicolumn{3}{c}{Frequency (Zipf)} & \multicolumn{3}{c}{\# Phonemes} & \multicolumn{3}{c}{Phon. similarity (Levenshtein)} \\ 
\cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
 & \emph{N} & Mean & \emph{SD} & Range & Mean & \emph{SD} & Range & Mean & \emph{SD} & Range \\ 
\midrule
\multicolumn{11}{l}{Catalan} \\ 
\midrule
A & 278 & $4.39$ & $0.63$ & $2.61$–$6.52$ & $5.28$ & $1.73$ & 2–14 & $37.32\%$ & $26.94\%$ & $0.00\%$–$100.00\%$ \\ 
B & 274 & $4.39$ & $0.69$ & $1.90$–$6.31$ & $5.20$ & $1.77$ & 2–14 & $35.96\%$ & $24.74\%$ & $0.00\%$–$100.00\%$ \\ 
C & 270 & $4.36$ & $0.65$ & $2.49$–$6.54$ & $5.46$ & $1.88$ & 2–14 & $36.51\%$ & $26.11\%$ & $0.00\%$–$100.00\%$ \\ 
D & 269 & $4.38$ & $0.61$ & $2.10$–$6.10$ & $5.29$ & $1.70$ & 2–14 & $36.33\%$ & $26.40\%$ & $0.00\%$–$100.00\%$ \\ 
\midrule
\multicolumn{11}{l}{Spanish} \\ 
\midrule
A & 277 & $4.25$ & $0.74$ & $2.28$–$6.76$ & $5.66$ & $1.63$ & 3–14 & $37.33\%$ & $26.89\%$ & $0.00\%$–$100.00\%$ \\ 
B & 281 & $4.28$ & $0.79$ & $2.16$–$6.14$ & $5.67$ & $1.68$ & 2–14 & $36.18\%$ & $24.95\%$ & $0.00\%$–$100.00\%$ \\ 
C & 278 & $4.24$ & $0.77$ & $1.68$–$6.26$ & $5.79$ & $1.79$ & 3–14 & $36.53\%$ & $26.24\%$ & $0.00\%$–$100.00\%$ \\ 
D & 273 & $4.24$ & $0.72$ & $1.98$–$6.14$ & $5.72$ & $1.68$ & 2–14 & $36.59\%$ & $26.55\%$ & $0.00\%$–$100.00\%$ \\ 
\bottomrule
\end{longtable}

\hypertarget{analysis}{%
\subsection{Data analysis}\label{analysis}}

We gathered 258,095 observations, with each observation corresponding to
a single response of one participant to a given item in the
questionnaire). These observations correspond to 395 distinct
participants (of which 343 participated once, 33 twice, 14 three times,
and 5 four times), responding to 1199 distinct items (597 in Catalan,
602 in Spanish, 602 translation equivalents). Translation equivalents
(TEs) received a median of 236 responses (\emph{Min} = 118, \emph{Max} =
942), both languages summed together.

We modelled the probability of answering each response category
(\emph{No} \textless{} \emph{Understands} \textless{} \emph{Understands
and Says}) using a Bayesian, mixed-effects ordinal regression model.
This allowed us to simultaneously estimate the impact of item- and
participant-level predictors on the probability of word comprehension
and production. We included \(Age\), \(Frequency\), \(Phonemes\),
\(DoE\), and \(Levenshtein\) as fixed effects. We also included the
\(Age \times DoE \times Levenshtein\) three-way interaction, and its
corresponding two-way interactions. We added participants and items as
crossed random effects including all intercepts and slopes when
appropriate (Barr et al., 2013).

>>>>

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

\setstretch{1.25}
\hypertarget{sec-introduction}{%
\section{Introduction}\label{sec-introduction}}

\hypertarget{word-acquisition-during-toddlerhood-the-effect-of-language-experience}{%
\subsection{Word acquisition during toddlerhood: the effect of language
experience}\label{word-acquisition-during-toddlerhood-the-effect-of-language-experience}}

The foundations of word learning are in place early in age: before the
end of their first year of life infants start directing their gaze to
some objects when hearing their labels, according to both experimental
data (Bergelson \& Swingley, 2012, 2015; Jusczyk \& Aslin, 1995; Tincoff
\& Jusczyk, 1999) and parental reports (e.g., Fenson et al., 2007).
During the last half of their second year, they acquire new words at an
increasingly fast rate (Bergelson, 2020; Bloom, 2002; Fenson et al.,
1994; Goldfield \& Reznick, 1990; Mayor \& Plunkett, 2011). These early
stages of lexical acquisition are characterised by substantial variation
across children, reflected, for instance, on the variation on the number
of words they know (i.e.~vocabulary size, Fenson et al., 1994; Frank et
al., 2021) or on the proportion of those words that fall into the
category of nouns, as opposed to verbs, adjectives, or function words
Bates et al. (1994).

@@

Previous findings point to bilingual toddlers knowing, on average, less
words in each of their languages than monolinguals, but also to both
groups knowing a similar amount of words when the two languages are
aggregated. For example, Hoff et al. (2012) found that English-Spanish
bilingual toddlers in South Florida knew less words in English than
monolinguals, who only learnt English. Both groups knew a similar total
amount of words when both English and Spanish vocabularies were counted
together, highlighting the importance of collecting data on both
languages when assessing bilinguals' communicative development. Other
studies have provided converging evidence that bilinguals know a
similar--or even larger--number of words than monolinguals, only when
the languages are aggregated (Gonzalez-Barrero et al., 2020; Oller \&
Eilers, 2002; J. Patterson et al., 2004; J. L. Patterson, 2004; Pearson
et al., 1993; Pearson \& Fernández, 1994; Petitto et al., 2001; Smithson
et al., 2014). While these studies have mostly relied on samples of
bilingual children learning two relatively distant languages, as it is
the case of English and Spanish, it is unclear whether children learning
typologically more similar languages also know less words in each of
their languages than monolinguals. What role could linguistic distance
play during early vocabulary growth?

\hypertarget{language-distance-as-a-mediator-of-bilingual-vocabulary-growth}{%
\subsection{Language distance as a mediator of bilingual vocabulary
growth}\label{language-distance-as-a-mediator-of-bilingual-vocabulary-growth}}

@@

\hypertarget{a-candidate-account-lexical-non-selectiviy-and-parallel-activation}{%
\subsection{A candidate account: lexical non-selectiviy and parallel
activation}\label{a-candidate-account-lexical-non-selectiviy-and-parallel-activation}}

Floccia et al.~pointed to \emph{parallel activation} as the main
phenomenon underpinning their results. The parallel activation
hypothesis suggests that bilinguals activate both languages
simultaneously during speech production or comprehension, and that this
phenomenon is the result of the activation of lexical representations in
both languages, even when only one is in use during production (Costa et
al., 2000; Hoshino \& Kroll, 2008) or comprehension (Spivey \& Marian,
1999; Thierry \& Wu, 2007). One of the clearest examples of parallel
activation was provided by Costa et al. (2000). In this study,
Catalan-Spanish monolingual and bilingual adults were asked to name
pictures of common objects in Spanish. In half of the trials, the object
labels were cognates in Spanish and Catalan (\emph{árbol}-\emph{arbre},
translations of \emph{tree}), whereas in the other half of the trials
labels were non-cognates (\emph{mesa}-\emph{taula}, translations of
\emph{table}). Bilinguals named cognate pictures faster than non-cognate
pictures, even after adjusting for the lexical frequency of the items.
Importantly, Spanish monolinguals--unfamiliar with the Catalan
translations of the Spanish words they uttered--did not show this
effect. These results suggest that bilinguals' Catalan phonology was
activated during the production of Spanish words, facilitating the
naming of cognate pictures. Several subsequent studies have also
provided similar evidence in comprehension in children (Poulin-Dubois et
al., 2013; e.g., Von Holzen \& Mani, 2012). Parallel activation is
therefore a plausible explanation for Floccia et al.'s results:
cognateness increases the amount of cross-language activation,
facilitating word acquisition, and ultimately leading to children
learning language pairs with a larger proportion of cognates are
predicted to show larger vocabulary sizes. This account is is line with
previous studies suggesting that the acquisition of new words is
facilitated by their phonological or semantic similarity with other
words, already acquired or not (Fourtassi et al., 2020; e.g., Hills et
al., 2009; Jones \& Brandt, 2019; Laing, 2022). Importantly, bilinguals
seem to be more likely to acquire words for which their translation
equivalent has been already learnt in the other language, suggesting
that semantic similarity also facilitates word acquisition across
languages (Bilson et al., 2015).

If phonological similarity also plays a cross-language facilitation role
during word acquisition, cognate translation equivalents--which share
both semantic and phonological similarity--should be acquired earlier
that non-cognate translation equivalents--which share semantic
similarity, but not phonological similarity. Evidence supporting and
earlier age of acquisition for cognates is scarce. Bosch \& Ramon-Casas
(2014) used vocabulary parental reports (152 lexical items) from 48
Catalan-Spanish bilinguals aged 18 months, and found that cognates
represented a larger proportion of participant's vocabulary than
non-cognates. Schelletter (2002) reported a longitudinal single case of
one English-German bilingual who produced cognates earlier than
non-cognates, on average. The low sample size in these two studies makes
it challenging to draw strong conclusions about the effect of
cognateness on vocabulary growth. On the other hand, Floccia et al.'s
estimates are statically more reliable given their (much larger) sample
size, but their study was not aimed at testing the effect of cognateness
on age of acquisition directly. In their discussion the authors state
the following (pp.~70):

@@

\hypertarget{two-competing-mechanisms-for-a-cognate-facilitation-during-word-acquisition}{%
\subsection{Two competing mechanisms for a cognate facilitation during
word
acquisition}\label{two-competing-mechanisms-for-a-cognate-facilitation-during-word-acquisition}}

As Mitchell et al. (2022) note in their discussion, the facilitative
effect of cognateness on word acquisition can be explained by at least
two alternative but arguably incompatible mechanisms. On the one hand,
cognateness might operate at the pre-lexical level as a bidirectional
facilitator of the acquisition of the two word-forms of a translation
pair: every exposure to either word-form activates the other, and
therefore strengthens the association between both word-forms and their
referent with every exposure. For instance, hearing \emph{flower} in
English might activate \emph{floeur} in French, and \emph{vice versa}.
Repeated exposure to the English or the French words eventually leads to
an earlier acquisition for both of them, as compared to non-cognate
translation equivalents like \emph{cow} and \emph{vache}, which do not
activate each other via phonological similarity. If this is the case,
then cognateness should increase the probability of acquisition of the
word-forms in both languages equally.

Another possibility is that cognateness plays a role in the acquisition
of a translation pair only when at least one of the word-forms has been
acquired. This account departs from the premise that phonological
similarity operates at the lexical level to facilitate word learning
across languages: if the child has still not formed a lexical
representation with sufficient phonological detail for one of the word
forms, the acquisition of its translation in the other language cannot
be facilitated by cognateness. This implies that, if cognateness
facilitates the acquisition of a word form, it can only do so after the
child has acquired one of the word forms of the translation pair. Given
that children are more likely to acquire words from languages to which
they are exposed more often (Cattani et al., 2014; David \& Wei, 2008;
Thordardottir et al., 2006), the acquisition of words in the language of
lower exposure should, on average, be more susceptible to the effect of
cognateness.

In line with this second hypothesis, the size of the effect of
linguistic similarity on vocabulary size that Floccia et al.~reported
was larger in the additional language (language of lower exposure)
vocabulary than in English vocabulary (language of higher exposure. Most
participants in their sample were English-dominant, meaning that their
relative amount of exposure to English was larger than in the additional
language. Therefore, participants may have, on average, learnt the
English word-form of translation equivalents earlier than the word-form
in the additional language. If this is the case, then the acquisition of
English words by English-dominant participants would rarely benefit from
their cognate status (the other word-form is not available yet), while
the acquisition of words in the additional language would benefit from
their phonological similarity with the (available) English form.

\hypertarget{the-present-study}{%
\subsection{The present study}\label{the-present-study}}

In this study, we investigated the role of cognateness on the
acquisition of translation equivalents, and if such effect is
conditional to the amount of time the child is exposed to the
corresponding language of each member of the translation pair. Using an
online vocabulary checklist--designed specifically for this study--we
collected data from a sample of children aged 10 to 35 months learning
Catalan and/or Spanish, with varying degrees of exposure to each
language. We then adopted an Bayesian item response theory (\emph{IRT})
approach to model the probability of participants being reported by
their parents to understand or understand \emph{and} say each word in
the checklist, conditional to its cognate status in Catalan and Spanish
and participants' degree of exposure to the corresponding language,
while adjusting for other indices of word difficulty--lexical frequency
(Goodman et al., 2008) and length (phonemes)-- and participant
skill--age (see Kachergis et al., 2022 for a similar approach).

\hypertarget{sec-methods}{%
\section{Methods}\label{sec-methods}}

\hypertarget{sec-participants}{%
\subsection{Participants}\label{sec-participants}}

@@

\hypertarget{sec-questionnaire}{%
\subsection{Questionnaire}\label{sec-questionnaire}}

To collect vocabulary data from participants, we created an \emph{ad
hoc} questionnaire, MultiLex (Garcia-Castro, 2022). This questionnaire
was inspired in the MacArthur-Bates Communicative Development Inventory
(Fenson et al., 2007) and its adaptations to other languages, and was
implemented on-line using the formR platform (Arslan et al., 2020). This
questionnaire is structured in three blocks: a (1) language
questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish
vocabulary checklists. Vocabulary checklists followed a similar
structure as the Oxford Communicative Developmental Inventory (Hamilton
et al., 2000) and consisted in two lists of words: one in Catalan and
one in Spanish. The Catalan inventory contained 798 items and the
Spanish inventory contained 802. Items in one language were translation
equivalents of the items in the other language (e.g., whenever
\emph{gos} {[}dog{]} was included in the Catalan inventory, the word
\emph{perro} was included in the Spanish inventory), roughly following a
one-to-one mapping. In case two translation equivalents were possible
for a given word, both were included as separate items (e.g., Catalan
\emph{acabar} {[}\emph{to finish}{]} and Spanish \emph{acabar} and
\emph{terminar}), or merged them into a single item (e.g., Spanish
\emph{mono} {[}\emph{monkey}{]} and Catalan \emph{mono/mico}). We
included items from a diverse sample of 27 semantic/functional
categories.

For each word in the vocabulary checklists, we asked parents to report
whether their child was able to understand it, understand \emph{and} say
it, or did not understand or say it (checked out by default). Some
families filled a long version of the vocabulary checklists (803
translation equivalents; 798 items in Catalan, 802 items in Spanish),
while others filled a shorter version (\textasciitilde400 translation
equivalents, \textasciitilde400 items in Catalan, \textasciitilde400
items in Spanish). These last families were randomly allocated into one
of four different subsets of the complete list of items. These lists
were designed so that each contained a representative sub-sample of the
items from the complete list. Semantic/functional categories with less
than 16 items--thus resulting in less than four items after dividing it
in four lists--were not divided in the short version of the
questionnaire: all of their items were included in the four lists.
Another subset of items that were part of the trial lists of some
experiments in the lab were also included in all versions. For the
analyses in the present study, we considered responses to words
corresponding to nouns, verbs, and adjectives (Fourtassi et al., 2020).
We excluded multi-word items (e.g., \emph{barrita de cereales} {[}cereal
bar{]}) and items that included more than one word-form (e.g.,
\emph{mono/mico}).

\hypertarget{variables-of-interest}{%
\subsection{Variables of interest}\label{variables-of-interest}}

We extracted four variables of interest for each word-form. (1) Word
lexical frequency (\(Frequency\)): lexical frequency of the word in its
corresponding language, expressed as Zipf scores (Van Heuven et al.,
2014; Zipf, 1949). This variable ranges from 0 to 7, and follows and
approximates a normal distribution, with most values in a corpus ranging
from 3 to 5 points. Lexical frequencies were extracted from an adult
subtitle-based database: SUBTLEX-CAT (Boada et al., 2020) for Catalan
words and SUBTLEX-ESP (Cuetos et al., 2012) for Spanish words. Using
child-based estimates of lexical frequency like CHILDES was not possible
due to the low number of Catalan participants and tokens in the
available corpora. Responses to words with missing lexical frequency
score were excluded from analyses (1.22\% of the total number of items).
(2) Word length (\(Phonemes\)): number of phonemes in the phonological
transcription of the word-form in International Phonological association
format. (3) Phonological similarity (\(Levenshtein\)): normalised
Levenshtein similarity between the word-form and its translation in the
other language (Levenshtein et al., 1966). This score is calculated by
first calculating the Levenshtein distance between the two
transcriptions (number of insertions, deletions or replacements needed
for the shortest transcription to become identical to the longer
transcription), then dividing the resulting value by the length of the
longest transcription, and finally subtracting this value from 1. This
results in a proportion that indicates how much the two phonological
transcriptions of the translation equivalent are similar to each other,
ranging from 0\% (no similarity at all) to 100\% (both transcriptions
are identical) (see Floccia et al., 2018; Fourtassi et al., 2020; Laing,
2022 for similar approaches). (4) Participant age: number of months
elapsed between participants' birth date and questionnaire completion.
Finally, (5) participant degree of exposure to the words' language
(\(DoE\)): percentage of exposure to the language the word belonged to.
For example, for a participant with 90\% exposure to Catalan, and 10\%
to Spanish, the DoE of the Catalan word \emph{taula} would be 90\%, and
the DoE for the Spanish word \emph{mesa} would be 10\%.
Table~\ref{tbl-item-summary} shows a summary of the distribution of the
word-level predictors across languages in and questionnaire versions.

\hypertarget{tbl-item-summary}{}
\begin{longtable}{lrrrrrrrrrr}
\caption{\label{tbl-item-summary}Lexical properties of the word-forms and translation pairs included in
the vocabulary checklist and our analyses. We report the number of
items, mean, standard deviation (SD), and range of the lexical frequency
(epxressed in Zipf scores), length (number of phonemes) of the word
forms, and the phonological similarity (normalised Levenshtein
similarity) with their translation equivalent. Statistics are reported
separately for Catalan and Spanish, and for the different vocabulary
checklists that participants were allocated to (A, B, C, D). }\tabularnewline

\toprule
 &  & \multicolumn{3}{c}{Frequency (Zipf)} & \multicolumn{3}{c}{Length (Phonemes)} & \multicolumn{3}{c}{Phon. similarity (Levenshtein)} \\ 
\cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
 & \emph{N} & Mean & \emph{SD} & Range & Mean & \emph{SD} & Range & Mean & \emph{SD} & Range \\ 
\midrule
\multicolumn{11}{l}{Catalan} \\ 
\midrule
A & 278 & $4.39$ & $0.63$ & $2.61$–$6.52$ & $5.28$ & $1.73$ & 2–14 & $37.32\%$ & $26.94\%$ & $0.00\%$–$100.00\%$ \\ 
B & 274 & $4.39$ & $0.69$ & $1.90$–$6.31$ & $5.20$ & $1.77$ & 2–14 & $35.96\%$ & $24.74\%$ & $0.00\%$–$100.00\%$ \\ 
C & 270 & $4.36$ & $0.65$ & $2.49$–$6.54$ & $5.46$ & $1.88$ & 2–14 & $36.51\%$ & $26.11\%$ & $0.00\%$–$100.00\%$ \\ 
D & 269 & $4.38$ & $0.61$ & $2.10$–$6.10$ & $5.29$ & $1.70$ & 2–14 & $36.33\%$ & $26.40\%$ & $0.00\%$–$100.00\%$ \\ 
\midrule
\multicolumn{11}{l}{Spanish} \\ 
\midrule
A & 277 & $4.25$ & $0.74$ & $2.28$–$6.76$ & $5.66$ & $1.63$ & 3–14 & $37.33\%$ & $26.89\%$ & $0.00\%$–$100.00\%$ \\ 
B & 281 & $4.28$ & $0.79$ & $2.16$–$6.14$ & $5.67$ & $1.68$ & 2–14 & $36.18\%$ & $24.95\%$ & $0.00\%$–$100.00\%$ \\ 
C & 278 & $4.24$ & $0.77$ & $1.68$–$6.26$ & $5.79$ & $1.79$ & 3–14 & $36.53\%$ & $26.24\%$ & $0.00\%$–$100.00\%$ \\ 
D & 273 & $4.24$ & $0.72$ & $1.98$–$6.14$ & $5.72$ & $1.68$ & 2–14 & $36.59\%$ & $26.55\%$ & $0.00\%$–$100.00\%$ \\ 
\bottomrule
\end{longtable}

\hypertarget{sec-analysis}{%
\subsection{Data analysis}\label{sec-analysis}}

We gathered 258,095 observations, with each observation corresponding to
a single response of one participant to a given item in the
questionnaire). These observations correspond to 395 distinct
participants (of which 343 participated once, 33 twice, 14 three times,
and 5 four times), responding to 1,199 distinct items (597 in Catalan,
602 in Spanish, 602 translation equivalents). Translation equivalents
(TEs) received a median of 236 responses (\emph{Min} = 118, \emph{Max} =
942), both languages summed together.
